---
title: "DATA607 - Project 1 - Chess Tournament"
author: "Asher Meyers & Chris Martin"
output:
  html_document:
    highlight: kate
    theme: united
---

#Project   1 - Chess Tournament

###Loading the Data
First, load the *stringr* library and dataset *TournamentInfo*. To do this, we import the csv file (columns seperated by *|*) and will have to remove the empty lines filled with *-*'s.


```{r}
library("stringr")
library("knitr")
TournamentInfo <- read.csv(url("https://raw.githubusercontent.com/chrisgmartin/DATA607/master/tournamentinfo.txt"),
                           sep = "|", stringsAsFactors=FALSE,
                           col.names = c("Pair", "Name", "Total", "R1", "R2", "R3", "R4", "R5", "R6", "R7", ""), 
                           skip = 1)
TournamentInfo <- TournamentInfo[seq(-2,-194,-3),]
kable(head(TournamentInfo))
```

***

###Seperating Rows for Each Player
Since we have the tournament info imported, the table shows that each player is displayed in two rows: the first includes the player name, result (win, loss, draw, etc.), and opponent number; the second includes the player's state, player ID, rating (pre and post tournament), and some other various information. Since we're only looking for specific information it'll be important to merge and display the information we need. Here we can seperate these two rows for each player into two tables with one row for each player.


```{r}
JoinedRows <- data.frame(TournamentInfo[seq(2,2*194/3,2),],TournamentInfo[seq(3,2*194/3,2),])
kable(head(JoinedRows))
```

***

###Adding Pre-Tournament Ratings
####Gathering Pre-Tournament Ratings

Next we'll pull out the Pre-Tournament Rating from the *NoNamesRows* table. From the TournamentInfo table, you can see that the Pre-Tournament Rating is a 3-4 digit number in the *Player.Name* column preceeding the characters: *"R: "*. Using the *str_locate* function we can locate the point for each row where the rating appears and then use the *str_sub* function to extract the string in the that location. Finally, since these strings are integers (which will help us in further calculations) we will want to extract them as integers.

```{r}
PreRatingLoc <- c(str_locate(JoinedRows[,13],"R: ")[1,2]+1, str_locate(JoinedRows[,13], "R: ")[1,2]+4)
PreRatings <- as.integer(str_sub(JoinedRows[,13],PreRatingLoc[1],PreRatingLoc[2]))
kable(head(PreRatings))
```


####Merging the Named and Non-Named tables with Pre-Tournament Ratings
Now that we have our tables, we can start to combine them. We'll call this new table **ChessTable** and pull in the Player Name, State, Total Number of Points, and Pre-Rating.

```{r}
ChessTable <- data.frame(c(1:64), sapply(JoinedRows[2],str_trim),JoinedRows[12],JoinedRows[3], PreRatings)
colnames(ChessTable) <- c("Number", "Name", "State", "Points", "PreRating")
kable(head(ChessTable))
```

***

###Adding Average-Opponent Ratings
####Gathering Opponent List

The last piece we'll need to add is the average opponent's Pre-Tournament Rating. To get this we'll create another table called *OppRatings*, which will list each opponent the players will face in the tournament.


```{r}
Opponents <- data.frame(1, as.list(as.numeric(str_extract(JoinedRows[1,4:10],"[0-9]{1,2}"))))
colnames(Opponents) <- c("Number", "Opp1","Opp2","Opp3","Opp4","Opp5","Opp6","Opp7")
for (i in 2:64) {
  Opponents <- rbind(Opponents, as.numeric(str_extract(JoinedRows[i,4:10],"[0-9]{1,2}")))
}
Opponents$Number <- c(1:64)
kable(head(Opponents))
```


####Gathering Opponent Pre-Tournament Ratings

From the list of each players' opponents, we'll pull in the opponents' score, number of games played, and get an average opponent Pre-Tournament Rating.

```{r}
Opponents <- data.frame(c(1:64),
                        ChessTable$PreRating[Opponents$Opp1[1:64]],
                        ChessTable$PreRating[Opponents$Opp2[1:64]],
                        ChessTable$PreRating[Opponents$Opp3[1:64]],
                        ChessTable$PreRating[Opponents$Opp4[1:64]],
                        ChessTable$PreRating[Opponents$Opp5[1:64]],
                        ChessTable$PreRating[Opponents$Opp6[1:64]],
                        ChessTable$PreRating[Opponents$Opp7[1:64]])
Opponents$Average <- round(rowMeans(Opponents[2:8],na.rm=TRUE), digits=0)
colnames(Opponents) <- c("Number", "Opp1", "Opp2", "Opp3", "Opp4", "Opp5", "Opp6", "Opp7", "Average")
kable(head(Opponents))
```

***

###Merging all of the components into one table

```{r}
ChessTable$OpponentAverage <- Opponents$Average
kable(head(ChessTable))
```

###Plotting the table

```{r}
require(ggplot2)
ggplot(ChessTable, aes(x=PreRating, y=OpponentAverage, label = Name, colour = Points)) + geom_point() + geom_text(aes(label=ifelse(PreRating>1700, as.character(Name),'')), size = 2, hjust=0) + xlim(0,2000) + ggtitle("Player to Opponent Scatterplot")
```


***

###Predicting an eigth match win

This tournament just got out of hand: against the rules of the tournament, an eigth round against a random player in the tournament has been set! Let's find the chances that a player will face-off and win against their random opponent:

####Selecting the random opponent

We'll start by selecting (at random, without replacement) which opponent our player will face, and then predict the chances of a win.

```{r}
ChessTable$Opp8 <- sample(1:length(ChessTable[,1]), 64, replace=F)
ChessTable$Opp8Rating <- ChessTable$PreRating[ChessTable$Opp8[1:64]]
kable(head(ChessTable))
```

####Calculating the chance of winning

Using the calculation WP = 1 / [10^ (dR/400) + 1 ] where dR is the difference in ratings, we can come up with a Winning Probability[1]

```{r}
ChessTable$ChanceOfWin <- paste(round((1 / (10^ ((ChessTable$Opp8Rating - ChessTable$PreRating) / 400) + 1)) * 100, 2), "%", sep = "")
kable(head(ChessTable))
```


[1] Reference: http://www.caissa.com/support/chess-ratings.php